
\chapter{Acknowledgments}

In this last part I want to express my sincere thanks to all the people that helped, assisted and guided me for this project.\\

First of all I would like to thank my supervisor Prof. Dr. Benjamin Dietzek-Ivanšić whose attention to detail and ability to ask the right questions at critical moments significantly improved the quality of this research. Furthermore all the people of the plasma group at IOM and especially high praise has to be given to Dr. Dmitry Kalanov, who was there for every issue and every success always looking for a solution and deduction, acting like a mentor. Additionally he is such a warm and friendly collegue just like the next person, who I want to lay out my gratitude towards: Dr. Yeliz Unutulmazsoy believed in me from the beginning and did everything imaginable to make life at IOM easy and enjoyable. Subsequently I want to thank Ramadan Saleh and Uwe Gey with which I spent a lot of fun and quality time in the laboratory keeping the atmosphere light and productive. Furthermore I want to thank Jens Bauer who with his excellent expertise could provide me with EDX/SEM data for very thin films.\\

Secondly I want to thank my amazing friend group; including Visar who introduced me to IOM (and hopefully more ;-P); my roommates Xuewei and Tom who motivated me with loads of cuppas and food ; Berta for an awesome arm sleeve tattoo and hopefully the next part this year; Kylian, Vincenz and Oliver who were always there for the mandatory Counterstrike tilt; I want single out Dennis DJ netto whose music motivated me through the hard days and is just straight fire; Chandramouli and Thiyagarajan for introducing me to amazing dishes and providing true friendship; and finally Visar who is always there to go eat Toscka at Syrian bistro.\\

Thirdly I want to thank my supporting family for always being there for me in all moments of life. In keiner besonderen Reihenfolge sondern nur wie es mir gerade eingefallen ist: Anne als unser Dungeon Meister, was mit Theo und mir bestimmt nicht leicht ist und auch immer schwerer wird, Msma die immer an meinen Fortschritt interessiert ist und mich sogar im Labor besucht hat, Papa der mich an manchen Stellen vielleicht sogar zu viel preist und alle anderen Familienmitglieder, die mir stets ein Gefühl der Wertschätzung vermitteln.\\

Fourthly I want to mention a few things which just make life better, Sox, your scratches keep my arm and hands always scared but your purring makes it all worth it. My childhood obsession of star trek, which got me interested in science, the world and society, combating stereotypes and prejudice making me the man I am today. Finally websites like Jetpunk, Sporkle and Geoguessr, which always spark my need to learn more about Geography, History and General Knowledge.